\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology

### 2022-02-12 jsdse/Manuscript.tex
rewriting to approachable

\if0\blind
{
  \title{\bf Teaching analysis reproducibility with accessibility: Enriching secondary documentation to support skill level diversity of users.}
  \author{Luna L. Sanchez Reyes\thanks{
    The authors gratefully acknowledge ``Sustaining the Open Tree of Life``, NSF ABI No. 1759838, and ABI No. 1759846.}\hspace{.2cm}\\
    School of Natural Sciences, University of California, Merced\\
    and \\
    Emily Jane McTavish \\
    School of Natural Sciences, University of California, Merced}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Teaching analysis reproducibility with accessibility: Enriching secondary documentation to support skill level diversity of users.}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Research reproducibility is essential for scientific development. Yet, rates of reproducibility
are low, especially in the natural sciences. As increasingly more research is relying on
computing tools and software, efforts for improving
reproducibility rates have focused on digital availability of research workflows
as computer code, as well as raw and processed data in computer readable form.
However, making research products digitally available does not
necessarily ensures that
they can be easily reached by the majority of a target user base, let alone the general population,
making them effectively inaccessible, counteracting availability, and hindering reproducibility.
Thus, accessibility of research products is a key aspect for the long term improvement
of reproducibility rates.
Using an example within evolutionary biology, we identify inclusivity barriers to accessibility,
% accessibility typically refers to addressing differences in abilities (visual, hearing, cognitive).
research products in the natural sciences:
%% list the barriers
use of highly specialized computing jargon and programming techniques; high cognitive load,
unspecified, unclear or lengthy goals; content-focused descriptions instead of user-focused;
inflexible learning environment; unapproachable (cold or intimidating) language;
and little to no diversity of representation of information.

Then, we propose a set of principles to improve accessibility of code and data,
and illustrate their application in a set of teaching materials for evolutionary biology that we put
to work in the field.
Finally, we elaborate on the general application of these principles for
documenting scientific workflows, to provide present learners
and future researchers with tools for successful scientific reproducibility.
%> ** add a sentence on implications for teaching reproducibility

@@

%% **Introduce OpenTree in abstract??

@@

Research reproducibility --the extent to which consistent results are obtained when
a scientific experiment or research workflow is repeated \citep{repdef2021}--
is a key aspect of the advancement
of science, as it constitutes a minimum standard that allows understanding research products
(e.g., methods, data, analysis, results, etc.)
to determine their reliability and generality, and eventually build up scientific
knowledge and applications based on those products
\citep{king1995replication, peng2011reproducible, powers2019open}.
In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe},
which has elicited concerns about a crisis in the field \citep{baker2016reproducibility}.
%%% The scientific community has united to incentivize cultural changes that will improve
%%% reproducibility rates long term, such as transparency, availability, and workflow
%%% automatization, to name a few \citep{peng2015reproducibility}.
In response, the scientific community has been developing new principles and standards to incentivize
cultural changes in an effort to improve long term reproducibility rates in the natural sciences
\citep{peng2015reproducibility, wilkinson2016fair}.
A standard for reproducibility that has received much attention is scientific availability, which
we define as a property denoting that a research product can be reached (acquired, copied, analyzed,
processed and/or reused) at no financial, legal or technical cost \citep{arnold2019turing},
and without geographic, demographic, social or temporal barriers for the population \citep{fecher2014open}.
In this paper, we argue that availability can not be fully achieved without accessibility (Box 1).

We recognize that the term accessibility is used in relation to various aspects of
the human experience. In relation to digital resources, these aspects have been broadly
categorized in four axis: perception, operability, understanding and robustness \citep{chisholm2001web, caldwell2008web}
% (visual, motor, auditory, cognitive, social, psychological, educational, etc.).
Some forms of accessibility do not support automatization nor scalability.
For example, Graphical User Interfaces (GUIs) are accessible, but not scalable and rarely automatizable.

we focus on X types of accessibility

% not all types of accessibility support reproducibility
% while accessibility is key to support availability, and availability is ket for reproducibility,
% some forms of accessibility do not support reproducibility.

We identify factors that have particularly affected accessibility in
the natural sciences, and ways the scientific community can address them to create reproducible
research workflows. These principles can be generalized and integrated
into the undergraduate and graduate school STEM curriculum, either for courses specialized
in reproducibility or within other subject areas, as a necessary component of
successful and impactful science.

@@

\bigskip
\bigskip

\noindent\fbox{%
    \parbox{\textwidth}{%
    \textbf{Box 1. The role of accessibility in improving reproducibility.}
      % Availability is defined as the fact or possibility that something can be bought,
      % used, or reached by someone \citep{availability2022cambridge}.
      % Accessibility is defined as "the fact of being able to be reached or obtained easily"
      % \citep{accessibility2022cambridge}
      Availability and accessibility are conceptually accepted as synonyms for ``the fact
      of being able to be reached" \citep{available-accessible2022cambridge}.
      However, accessibility's definition goes further and describes it as the
      ``fact of being able to be reached \textbf{easily}" \citep{accessibility2022cambridge}.
      A secondary definition of accessibility elaborates more on this: ``the quality
      of being \textbf{easy to understand}''.
      % Yet, in practice, availability does not imply accessibility.

      What we realize from these definitions is that, in practice, something might
      be available but not necessarily accessible,
      hindering the realization of availability.
      To illustrate the practical difference between availability and accessibility we
      will use an allegory that goes like this:

      ** Change marshmallow story to cake

      ** Explain different types of barriers to accessibility within the example, then within science

      ** Explain why we focus on psychological/social/pedagogical barriers

      ** Gauging accessibility of a resource is tricky because it depends on an
      individual experience of what is in effect \textbf{easy} for someone.
      \textbf{Easy} is a subjective quality that fully depends on individual perceptions
      resulting from individual contexts.
      As a society we are moving towards validating and welcoming all human experiences.
      To improve accessibility levels in the natural sciences, it
      is important to account for the factors that are creating different experiences
      of what is \textbf{easy}.

      ** Now, explain different types of barriers to accessibility in science
      (Jessica: Maybe say that lack of accessibility can be due to a number of
      factors (physical, resources, economic), but one critical factor is the use
      of impenetrable expert language.)
    }%
}
\bigskip

\section*{An example from phylogenetics}
\label{sec:example}

To elaborate on our thesis,
%% we use a common scientific workflow in phylogenetics,
%% ** Jessica mentions that workflow is confusing as she thought it referred to the OpenTree
%% workflow, but we actually mean the more general workflow of reconstructing a phylogenetic tree.
%% She suggests something more general:
we choose an example from the research field of phylogenetics,
a discipline within evolutionary biology. Phylogenetics focuses
on investigating the history of shared ancestry of living organisms from biological data,
and represents
this evolutionary history in a diagram known as a phylogeny (Figure \ref{fig:tree}).
Phylogenies provide the
basis to study and understand all biological processes in an evolutionary context
\citep{dobzhansky1973nothing}. Hence,
% Given the importance of phylogenies for biological research,
it appears that improving reproducibility rates
in phylogenetics has the potential to positively impact research across the natural sciences.

@@

The Open Tree of Life project (OpenTree) has developed a computational platform that provides
digital availability of phylogenetic results from vetted, state-of-the-art research
in the field \citep{opentreeoflife2019synth}.
%% by creating a standardized database of phylogenetic data that is used to synthesize
%% a single phylogenetic tree encompassing all life
OpenTree's database, the Phylesystem \citep{mctavish2015phylesystem}, provides
usability of phylogenetic data by standardizing organism names across biological databases,
and supporting digital data obtention in a computer readable format.

While this is accessible through their Graphical User Interface,
It does not support reproducibility

@@

While learners in the natural sciences have been engaging independently with R
and Python programming languages
as they represent two of the most widely used programming languages in the sciences
% for their fast and seamless statistics and data analysis environment
\citep{baker2017scientific}, computer programming is not traditionally
a core skill formally taught to biologists and naturalists
\citep{sayres2018bioinformatics, wright2019the, williams2019barriers}.
Moreover, the main usage of the OpenTree API services hints that tools intended to
improve availability and reproducibility come at a cost for the potential users:
 an increased required baseline computational knowledge and skillset
in the field.
As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools},
higher baseline computational skills are required across all natural sciences.
% "Thus efforts to increase scientific reproducibility should consider the ubiquity of computers in research."
% In order for reproducible computational tools to be adopted for research, they
% need to be readily accessible for present and future researchers.
Thus, efforts to increase reproducibility rates in the natural sciences must consider
the specifics of scientific workflows that rely on usage of programming languages, such as
availability and accessibility of data and code \citep{peng2011reproducible, sandve2013ten, powers2019open}.
% However, the utility of data resources is limited by the technical challenges of accessing the data.
% In order to motivate reproducible research, that gap needs to be bridged.
% In this work we focus on improving accessibility of code examples and documentation.

>>>>

\if0\blind
{
  \title{\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology}
  \author{Luna L. Sanchez Reyes\thanks{
    The authors gratefully acknowledge ``Sustaining the Open Tree of Life``, NSF ABI No. 1759838, and ABI No. 1759846.}\hspace{.2cm}\\
    School of Natural Sciences, University of California, Merced\\
    and \\
    Emily Jane McTavish \\
    School of Natural Sciences, University of California, Merced}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Research reproducibility is essential for scientific development. Yet, rates of reproducibility
are low, especially in the natural sciences. As increasingly more research is relying on
computing tools and software, efforts for improving
reproducibility rates have focused on making available research workflows
as computer code, as well as raw and processed data in computer readable form.
However, research products that are digitally available are not necessarily
friendly for learners and interested parties with little to no %%% different levels of
experience in the field. This renders research products unapproachable, which
counteracts availability, and hinders reproducibility short and long term.
To improve long term adoption of reproducible workflows in research, they
need to be made approachable for learners, the researchers of the future.

Using an example within evolutionary biology, we identify aspects of research workflows
that make them unapproachable to the general audience:
use of highly specialized computing jargon and programming techniques;
high cognitive load;
unspecified, unclear or lengthy goals;
content-focused descriptions instead of user-focused;
inflexible learning environment;
unapproachable (cold or intimidating) language; and
little to no diversity of representation of information.
Then, we propose a set of principles to improve the unapproachable aspects of
research workflows, and illustrate their application in a case study from
evolutionary biology that we used as teaching material.
Finally, we elaborate on the general application of these principles for
documenting research workflows and products, to provide present learners
and future researchers with tools for successful scientific reproducibility.
%> ** add a sentence on implications for teaching reproducibility

@@

@@

Research reproducibility --the extent to which consistent results are obtained when
a scientific experiment or research workflow is repeated \citep{repdef2021}--
is a key aspect of the advancement
of science, as it constitutes a minimum standard that allows understanding research products
(e.g., methods, data, analysis, results, etc.)
to determine their reliability and generality, and eventually build up scientific
knowledge and applications based on those products
\citep{king1995replication, peng2011reproducible, powers2019open}.
In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe},
which has elicited concerns about a crisis in the field \citep{baker2016reproducibility}.

In response, the scientific community has been developing new principles and standards to incentivize
cultural changes in an effort to improve long term reproducibility rates in the natural sciences
\citep{peng2015reproducibility, wilkinson2016fair}.
A standard for reproducibility that has received much attention is availability, which
we define as a property denoting that a research product can be reached (acquired, copied, analyzed,
processed and/or reused) at no financial, legal or technical cost \citep{arnold2019turing},
and without geographic, demographic, social or temporal barriers for the population \citep{fecher2014open}.

%% In this paper, we argue that availability can not be fully achieved
%% without approachability (Box 1).
However, research products that are digitally available are not necessarily
friendly for learners and interested parties with different levels of
experience in the field. This renders research products unapproachable, which
counteracts availability, and hinders reproducibility short and long term.
To support long term adoption of reproducible practices in research, research
workflows need to be made approachable for learners, the researchers of the future.

To elaborate on our thesis, we develop a case study from the research field of
phylogenetics, a discipline within evolutionary biology.
This case study allowed us to identify barriers that have made research workflows
largely unapproachable to a general audience in the natural sciences.
Then, we propose some principles for researchers to address these barriers, to
create research workflows that are reproducible by a larger audience.

@@

The principles proposed here can be generalized and integrated into the undergraduate
and graduate school STEM curriculum, either for courses specialized in reproducibility
or within other subject areas, as a necessary component of successful and impactful science.

%% \bigskip
%% \bigskip
%%
%% \noindent\fbox{%
%%     \parbox{\textwidth}{%
%%     \textbf{Box 1. The role of accessibility in improving reproducibility.}
%%       % Availability is defined as the fact or possibility that something can be bought,
%%       % used, or reached by someone \citep{availability2022cambridge}.
%%       % Accessibility is defined as "the fact of being able to be reached or obtained easily"
%%       % \citep{accessibility2022cambridge}
%%       Availability and accessibility are conceptually accepted as synonyms for ``the fact
%%       of being able to be reached" \citep{available-accessible2022cambridge}.
%%       However, accessibility's definition goes further and describes it as the
%%       ``fact of being able to be reached \textbf{easily}" \citep{accessibility2022cambridge}.
%%       A secondary definition of accessibility elaborates more on this: ``the quality
%%       of being \textbf{easy to understand}''.
%%       % Yet, in practice, availability does not imply accessibility.
%%
%%       What we realize from these definitions is that, in practice, something might
%%       be available but not necessarily accessible,
%%       hindering the realization of availability.
%%       To illustrate the practical difference between availability and accessibility we
%%       will use an allegory that goes like this:
%%
%%       ** Change marshmallow story to cake
%%
%%       ** Explain different types of barriers to accessibility within the example, then within science
%%
%%       ** Explain why we focus on psychological/social/pedagogical barriers
%%
%%       ** Gauging accessibility of a resource is tricky because it depends on an
%%       individual experience of what is in effect \textbf{easy} for someone.
%%       \textbf{Easy} is a subjective quality that fully depends on individual perceptions
%%       resulting from individual contexts.
%%       As a society we are moving towards validating and welcoming all human experiences.
%%       To improve accessibility levels in the natural sciences, it
%%       is important to account for the factors that are creating different experiences
%%       of what is \textbf{easy}.
%%
%%       ** Now, explain different types of barriers to accessibility in science
%%       (Jessica: Maybe say that lack of accessibility can be due to a number of
%%       factors (physical, resources, economic), but one critical factor is the use
%%       of impenetrable expert language.)
%%     }%
%% }
%% \bigskip

\section*{A case study from phylogenetics}
\label{sec:case}

Phylogenetics is a key discipline within evolutionary biology \citep{dobzhansky1973nothing}.
It focuses on investigating the history of shared ancestry of living and extinct
organisms using biological data,
and represents this evolutionary history with a diagram known as a phylogeny
or phylogenetic tree (because it grows through time and appears to have branches; Figure \ref{fig:tree}).
Phylogenies provide the basis to study and understand all biological processes
in an evolutionary context \citep{dobzhansky1973nothing}
Hence, it appears that improving reproducibility rates in phylogenetics has the
potential to positively impact research across the natural sciences.

@@

The Open Tree of Life project (OpenTree) has developed a computational platform that provides
digital availability of phylogenetic results from vetted, state-of-the-art research
in the field \citep{opentreeoflife2019synth}.
%% by creating a standardized database of phylogenetic data that is used to synthesize
%% a single phylogenetic tree encompassing all life
OpenTree's database, the Phylesystem \citep{mctavish2015phylesystem}, stores
phylogenetic data in a reusable form by applying a single standard to name organisms
in the stored phylogenies \citep{rees2017automated}.
%% It also supports digital data obtention in a computer readable format.

While this is available through their Graphical User Interface,
It does not support reproducibility

@@

While learners in the natural sciences have been engaging independently with R
and Python programming languages
as they represent two of the most widely used programming languages in the sciences
% for their fast and seamless statistics and data analysis environment
\citep{baker2017scientific}, computer programming is not traditionally
a core skill formally taught to biologists and naturalists
\citep{sayres2018bioinformatics, wright2019the, williams2019barriers}.
Moreover, the main usage of the OpenTree API services hints that tools intended to
improve availability and reproducibility come at a cost for the potential users:
 an increased required baseline computational knowledge and skillset
in the field.
As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools},
higher baseline computational skills are required across all natural sciences.
% "Thus efforts to increase scientific reproducibility should consider the ubiquity of computers in research."

Thus, efforts to increase reproducibility rates in the natural sciences must consider
the specifics of scientific workflows that rely on usage of programming languages, such as
availability and accessibility of data and code \citep{peng2011reproducible, sandve2013ten, powers2019open}.
% However, the utility of data resources is limited by the technical challenges of accessing the data.
% In order to motivate reproducible research, that gap needs to be bridged.
% In this work we focus on improving accessibility of code examples and documentation.
